Template for indicator documentation

### 2023-04-19 docs/template.Rmd

_Author and date:_
<!-- Write your name here -->

```{r}
Sys.Date()
```

@@

### <!-- Text here -->
<!-- Text here. Note that there are two parallell systems for defining ecosystem properties - one Norwegian and one international (SEEA EA). See man/SEEA EA examples.pdf for help -->

@@

## Reference state and values
### Reference state
<!-- Text here -->

>>>>

_Author and date:_
<!-- Write your name, month and year -->

@@

### SEEA EA (UN standard)
<!-- Text here. Note that there are two parallell systems for defining ecosystem properties - one Norwegian and one international (SEEA EA). See man/SEEA EA examples.pdf for help -->

@@

## Reference condition and values
### Reference condition
<!-- Text here. Note that the reference condition is defined in Fagsystemet for økologisk tilstand and it is the same for all indicators and ecosystems. So you can give a short and generic answer here. -->

### 2023-04-19 index.qmd

@@

On this site you may view the nitty gritty details for how certain indicators for ecosystem condition for Norway has been calculated.

@@

>>>>

---
title: "Indicators for Ecosystem Condition in Norway"
author: "Anders L. Kolstad"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
url: https://ninanor.github.io/ecosystemCondition/index.html
#cover-image: images/Picture1.png
description: |
  This is a site for documenting the analyses related to the development and design of ecosystem condition indicators.
biblio-style: apalike
csl: chicago-fullnote-bibliography.csl
---

@@

On this site you may view the details for how certain indicators for ecosystem condition for Norway have been calculated.

@@

## Citation {.unnumbered}

To cite this book please use the following citation:

Kolstad, A.L., Topper, J. Grainger, M.J., Sandvik, H. Indicators for Ecosystem Condition in Norway. https://ninanor.github.io/ecosystemCondition/index.html.doi [TBA]

## License {.unnumbered}

The book is licensed under a [CC By 4.0 licence](https://creativecommons.org/licenses/by/4.0/) which means that you can share and adapt the material
for any purpose. However, you must give appropriate credit, provide a link to the license, and indicate if changes were made. You can not add additional restrictions on the material even when adapted.

### 2023-04-19 medianSummerTemperature.qmd

# Climate indicators explained {#climate-indicators-explained}

@@

```{r, echo=F}
Ecosystem <- "All" # e.g. "Skog og fjell"
Egenskap  <- "Abiotiske egenskaper" # e.g. "Primærproduksjon"
ECT       <- "Physical state characteristics" # e.g. "Structural state characteristic"
Contact   <- "Anders L. Kolstad" # e.g. "Anders Kolstad"
```

@@

This chapters describes the development of a workflow for generating or preparing indicators based on interpolated climate data from [SeNorge](https://senorge.no/). I describe the process using dummy data and also discuss alternate approaches and dead ends that occured in the development process. For a succinct workflow example for a specific indicator, see [Median summer temperature](#median-summer-temp).

@@

This table explains them in more detail

```{r,warning=FALSE, message=FALSE, fig.cap="Table explaining the available climate parameters in the data set."}
senorgelayers <- read_delim("data/senorgelayers.txt", 
    delim = "\t", escape_double = FALSE, 
    trim_ws = TRUE)
DT::datatable(senorgelayers)
```

Here is the content of one of these folders (first 10 entries)

```{r}
(files_tm <- list.files(paste0(path, "/rr_tm")))[1:10]
```

There are `r length(files_tm)` files, i.e. `r length(files_tm)` years of data, one file per year.

Importing (proxy) file:

```{r, warning=F}
tg_1957 <- stars::read_ncdf(paste0(path, "/rr_tm/seNorge2018_1957.nc"),
                            var="tg")
```

I know the CRS, so setting it manually. Although I cannot rule out 32633, I don't think it matters.

```{r}
st_crs(tg_1957) <- 25833
```

The data has three dimension

```{r}
dim(tg_1957)
```

Initially the data had four attributes, but I subsettet to only include *tg*.

```{r}
names(tg_1957)
```

`tg_1957` is a stars proxy object and [most commands](https://r-spatial.github.io/stars/articles/stars8.html) will not initiate any change until later, typically I write to file and all the lazy operations are done in squeal. Therefore, I will prepare a test data set, which is smaller and which I can import to memory. Then I can perform the operations on that data set and we can see the results.

#### Dummy data

This test data is included in the {stars} package

```{r}
tif = system.file("tif/L7_ETMs.tif", package = "stars")
t1 = as.Date("1970-05-31")
x = read_stars(c(tif, tif, tif, tif), along = 
                  list(time = c(t1, t1+1, t1+50, t1+100)), 
               RasterIO = list(nXOff = c(1), 
                               nYOff = c(1), 
                               nXSize = 50, 
                               nYSize = 50, 
                               bands = c(1:6)))
```

A single attribute

```{r}
names(x)
```

I can rename it like this

```{r}
x <- setNames(x, "Attribute_A")
```

And I can add another dummy attribute.

```{r}
x <- x %>%
  mutate("Attribute_B" = Attribute_A/2)
```

The dummy data also has four dimensions

```{r}
dim(x)
```

X and y area the coordinates. Band is an integer:

```{r}
st_get_dimension_values(x, "band")
```

I will remove the *band* dimension.

```{r}
x <- x %>% filter(band==1) %>% adrop()
```

Time is four dates covering four months in 1970:

```{r}
(oldTime <- st_get_dimension_values(x, "time"))
```

I'll add some extra seasonal variation between to the mix

```{r}
x$Attribute_A[,,2] <- x$Attribute_A[,,1]*1.2
x$Attribute_A[,,3] <- x$Attribute_A[,,1]*1.4
x$Attribute_A[,,4] <- x$Attribute_A[,,1]*1.1
```

Now I create another four copies of this data, adding some random noise and a continuous decreasing trend.

```{r}
# Function to add random noise
myRandom <- function(x) x*rnorm(1,1,.05)
```

```{r}
y1 <- x
y2 <- x %>% st_apply(1:3, myRandom) %>% st_set_dimensions("time", values = oldTime %m+% years(-1)) %>% mutate(Attribute_A = Attribute_A*0.95)
y3 <- x %>% st_apply(1:3, myRandom) %>% st_set_dimensions("time", values = oldTime %m+% years(-2)) %>% mutate(Attribute_A = Attribute_A*0.90)
y4 <- x %>% st_apply(1:3, myRandom) %>% st_set_dimensions("time", values = oldTime %m+% years(-3)) %>% mutate(Attribute_A = Attribute_A*0.85)
y5 <- x %>% st_apply(1:3, myRandom) %>% st_set_dimensions("time", values = oldTime %m+% years(-4)) %>% mutate(Attribute_A = Attribute_A*0.80)

```{r}
# We combine the data into one cube for plotting:
temp <- c(y1, y2, y3, y4, y5)

ggplot() + 
  geom_stars(data = temp) +
  coord_equal() +
  theme_void() +
  scale_fill_viridis_c(option = "D") +  
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0))+
  facet_wrap(~time, ncol=4)
```

Saving these to file.

```{r, eval=F}
path <- "P:/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/dummy_data"
saveRDS(y1, paste0(path, "/y1.rds"))
saveRDS(y2, paste0(path, "/y2.rds"))
saveRDS(y3, paste0(path, "/y3.rds"))
saveRDS(y4, paste0(path, "/y4.rds"))
saveRDS(y5, paste0(path, "/y5.rds"))
```

##### Dummy regions and ET map

Here is some dummy data for accounting areas (regions) and ecosystem occurrences as well, in case I need them later.

```{r, fig.cap="Showing the example account area delineation (raster format)"}
dummy_aa <- x %>%
  filter(time == oldTime[1]) %>%
  adrop() %>%
  mutate(accountingAreas = rep(c(1,2), each=length(Attribute_A)/2)) %>%
  select(accountingAreas)

tm_shape(dummy_aa) +
  tm_raster(style="cat")
```

```{r, fig.cap="Showing the example Ecosystem type delineation data."}
dummy_et <- x %>%
  filter(time == oldTime[1]) %>%
  adrop() %>%
  mutate(ecosystemType = rep(c(1,NA), length.out = length(Attribute_A))) %>%
  select(ecosystemType)

tm_shape(dummy_et) +
  tm_raster(style="cat")
```

@@

```{r}
reg <- st_intersection(reg, nor)
```

```{r, fig.cap="Five accounting areas (regions) in Norway."}
tm_shape(reg) +
  tm_polygons(col="region")
```

@@

6.  Make trend figure and spatially aggregated maps,

### Collate variable data series

I include a for loop for importing the .nc files at the end.

### Filter data by dates

Say we want to calculate the mean summer temperature. We then want to exclude the data for the times that is not withing our definition of summer.

Let's try it on the dummy data. Remember this data had four time steps.

```{r}
st_get_dimension_values(x, "time")
```

Our real data has 365 time steps for each year.

First I define my start and end dates. I want to keep only the summer data, defined as jun - aug.

```{r}
start_month <-  "06"
end_month <- "08"
```

Then for each iteration I need to get the year as well

```{r}
(year_temp <- year(st_get_dimension_values(x, "time")[1]))
```

Then I can create a time interval object

```{r}
start <- ym(paste(year_temp, start_month, sep="-"))
end <-  ym(paste(year_temp, end_month, sep="-"))
(myInterval <- interval(start, end))
```

Then I filter

```{r}
x_aug <- x %>%
  select(Attribute_A) %>%
  adrop() %>%
  filter(time %within% myInterval)
st_get_dimension_values(x_aug, "time")
```

*Later I discover that \`dplyr::filter\` work after reading the whole object to file, and therefor it is too slow for our use. I therefore filter data a different way further down.*

### Aggregate across time within a year

For this example I want to calculate the mean temperature over the summer. I therefore need to aggregate over time. Many climate indicators will need this functionality. There are two ways, either using `st_apply` or using `aggregate`.

```{r}
temp_names <- year(st_get_dimension_values(x_aug, "time")[1])

microOut <- microbenchmark(
st_apply =  x_aug %>%
  st_apply(1:2, mean) %>%
  setNames(paste0("v_",temp_names)),
aggregate = x_aug %>%
  aggregate(by = "year", mean),
times=30
)
```

The time dimension is now gone as we have aggregated across it.

```{r, fig.cap="Comparing computation tome for two spatial aggregation functions."}
autoplot(microOut)
```

`st_apply` is slightly faster, but `aggregate` has the advantage of that it retains the time dimension, whereas for `st_apply` I need to set the attribute name to be the year. I will try to use `st_apply`.

But let me see if I can save time by masking to ET before doing this operation.

```{r, fig.cap="Demonstrating the effect of maskig the dummy data using a perfectly aligne raster ET map."}
x_masked <- x_aug
x_masked[is.na(dummy_et)] <- NA
plot(x_masked[,,,1])
```

```{r withMasking, fig.cap="Comparing computation time before and after masking of raster data. There is no performance increase by masking before aggregating."}
autoplot(
  microbenchmark(
No_masking =  x_aug %>%
  st_apply(1:2, mean) %>%
  setNames(paste0("v_",temp_names)),
Masking =  x_masked %>%
  st_apply(1:2, mean) %>%
  setNames(paste0("v_",temp_names))
 ), times=30
)

Since there is no increase in performance from masking, as we see in Fig. \@ref(fig: withMasking) then it basically means we it is slower to mask beforehand, since the masking was not past of the benchmark assessment.

So, this is my solution:

```{r}
# setNames dont work on stars proxy obejct, so I use rename instead
lookup <- setNames("mean", paste0("v_", temp_names))

x_summerMean <-   x_aug %>%
  st_apply(1:2, mean) %>%
  rename(all_of(lookup))
```

The *v\_* follows the naming convention we have developed for ourselves.

```{r, fig.cap="Plotting the dummy data showing Attribute_A for the two dates as small maps, and a larger map showing the mean for year 2018"}
ggarrange(
ggplot() + 
  geom_stars(data = x_aug) +
  coord_equal() +
  facet_wrap(~time) +
  theme_void() +
  scale_fill_viridis_c(option = "D") +  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)),

ggplot() + 
  geom_stars(data = x_summerMean) +
  coord_equal() +
  #facet_wrap(~time) +
  theme_void() +
  scale_fill_viridis_c(option = "D") +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0))
)
```

Here's the whole first step of the conceptual workflow in action, on the dummy data.

```{r}

# Setting up the parameters
dir <- substr(getwd(), 1,2)
path <- ifelse(dir == "C:",
       "P:/", 
       "/data/P-Prosjekter2/")
path2 <- paste0(path, 
                "41201785_okologisk_tilstand_2022_2023/data/climate_indicators/dummy_data")
start_month <-  "06"
end_month <- "08"
myFiles <- list.files(path2, pattern=".rds", full.names = T)
summerTemp_fullSeries <- NULL

# Looping though the files in the directory
system.time({
for(i in 1:length(myFiles)){

  temp <- readRDS(myFiles[i]) %>%
    select(Attribute_A)

  year_temp <- year(st_get_dimension_values(temp, "time")[1])
  start <- ym(paste(year_temp, start_month, sep="-"))
  end <-  ym(paste(year_temp, end_month, sep="-"))
  myInterval <- interval(start, end)
  lookup <- setNames("mean", paste0("v_", year_temp))

  temp <- temp %>%
    filter(time %within% myInterval) %>%
    st_apply(1:2, mean) %>%
    rename(all_of(lookup))

  summerTemp_fullSeries <- c(temp, summerTemp_fullSeries)
  rm(temp) # same computation time with and without this function, but more tidy this way
}

# Turn the attributes into a dimension and rename the new attribute
summerTemp_fullSeries <- summerTemp_fullSeries %>%
  merge(name="Year") %>%
  setNames("Attribute_A")
})
```

And this is the result.

```{r, fig.cap="Temporal aggregation on the dummy data set."}
ggplot()+
  geom_stars(data = summerTemp_fullSeries) +
  coord_equal() +
  theme_void() +
  scale_fill_viridis_c(option = "D") +  
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0))+
  facet_wrap(~Year)
```

Now let's try it on the real data.

*I initially tried to read stars proxy object and add steps to the call_list and execute these with `st_as_stars`, but this still returned proxy objects. Therefore I read the whole file in without proxy. To save time I can subset based on Julian dates.*

I will use [parallelization](https://tmieno2.github.io/R-as-GIS-for-Economists/EE.html) to speed things up.

@@

This takes about 20 sec per file/year, or 22 min on total. That is not too bad. About 6000 raster are read into memory. Here's a test for the effect of splitting over more cores.

```{r, eval=F}
cl <- makeCluster(2L)
cl2 <- makeCluster(6L)
cl3 <- makeCluster(10L)

tic("No cluster")
temp %>%
    #filter(time %within% myInterval) %>%
    st_apply(1:2, mean) %>%
    rename(all_of(lookup)) 
toc()

tic("Two clusters")
temp %>%
    #filter(time %within% myInterval) %>%
    st_apply(1:2, mean, CLUSTER = cl) %>%
    rename(all_of(lookup)) 
stopCluster(cl)
toc()

tic("Six clusters")
temp %>%
    #filter(time %within% myInterval) %>%
    st_apply(1:2, mean, CLUSTER = cl2) %>%
    rename(all_of(lookup)) 
stopCluster(cl2)
toc()

tic("Ten clusters")
temp %>%
    #filter(time %within% myInterval) %>%
    st_apply(1:2, mean, CLUSTER = cl3) %>%
    rename(all_of(lookup)) 
stopCluster(cl3)
toc()

-   No cluster: 21.401 sec elapsed

-   Two clusters: 21.546 sec elapsed

-   Six clusters: 14.548 sec elapsed

-   Ten clusters: 15.29 sec elapsed

-   Ten clusters (second run): 13.617 sec elapsed

The RStudio server has 48 cores. More parallel cores is probably on average faster. The NINA guidelines is to use max 10 cores, and to remember to close parallel cluster when done.

```{r, eval=F}
#saveRDS(real_temp_summer, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/real_temp_summer.rds")
#write_stars(real_temp_summer, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/real_temp_summer.tiff")
```

These files are about 500 MB, but the tiff file reads in a lot quicker then the rds file, so from now on I'll just export as tiff. *(This could perhaps have been solved with a `gc()`. Later I tried saving as RData with no speed issues problems, but then I had a problem with RData files seemingly becoming corrupted over time - stars objects saved as RData files would not bind together using `c.stars()` with other matching stars object and I think this could be related to a time stamp in the RData format. So, stick to tiff).*

```{r}
summer_median_temp <- read_stars(paste0(pData, "climate_indicators/aggregated_climate_time_series/real_temp_summer.tiff"),
                                 proxy=F)
```

```{r, eval=T}
names(summer_median_temp)
```

```{r, eval=T}
dim(summer_median_temp)
```

```{r, eval=T}
st_get_dimension_values(summer_median_temp, "Year")
```

Note that GTiff automatically renames the third dimension *band* and also renames the attribute.

```{r}
st_get_dimension_values(summer_median_temp, "band")
```

I can rename them.

```{r}
summer_median_temp <- summer_median_temp %>% 
  st_set_dimensions(names = c("x", "y", "v_YEAR")) %>%
  setNames("temperature")
dim(summer_median_temp)
```

```{r, fig.cap="Showing three random slices of the year dimension."}
ggplot()+
  geom_stars(data = summer_median_temp[,,,c(1,11,22)], downsample = c(10, 10, 0)) +
  coord_equal() +
  theme_void() +
  scale_fill_viridis_c(option = "D") +  
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0))+
  facet_wrap(~v_YEAR)
```

### Calculate reference values

#### Aggregate across reference years

We need to first to define a reference period and then to subset our data `summer_median_temp`. We can practice on the dummy data `y`.

```{r}
dir <- substr(getwd(), 1,2)
path <- ifelse(dir == "C:",
       "P:/", 
       "/data/P-Prosjekter2/")
path2 <- paste0(path, 
                "41201785_okologisk_tilstand_2022_2023/data/climate_indicators/dummy_data")
myFiles <- list.files(path2, pattern=".rds", full.names = T)
y <- NULL

# Looping though the files in the directory
for(i in 1:length(myFiles)){
  temp <- readRDS(myFiles[i])
  y <- c(temp, y)
}
```

```{r}
st_get_dimension_values(y, "time")
```

```{r}
y_filtered <- y %>%
  select(Attribute_A) %>%
  filter(time %within% interval("1961-01-01", "1990-12-31"))
```

Now we aggregate across years, again using st_apply.

```{r}
y_ref <- y_filtered %>%
  st_apply(c("x", "y"), FUN =  median_sd)
```

```{r}
dim(y_ref)
```

```{r}
st_get_dimension_values(y_ref, "median_sd")
```

It's perhaps easier if median and sd are unique attributes, rather than levels of a dimension.

```{r}
y_ref <- y_ref %>% split("median_sd")
```

The attribute name *mean* we can change this to something more meaningful.

```{r}
y_ref <- setNames(y_ref, c("reference_upper", "sd"))
```

```{r, fig.cap="Showing the upper reference levels and the standard deviation from the dummy data set."}
tmap_arrange(
tm_shape(y_ref)+
  tm_raster("reference_upper")
,
tm_shape(y_ref)+
  tm_raster("sd",
            palette = "-viridis")
)
```

Let's transfer this over to the real data. First we need to rename our dimension values and turn them back into dates.

@@

This is also something I can export.

```{r, eval=F}
# Note that write_stars can only export a single attribute, but we have put the 'attributes' as dimensions, and these will be read in as 'band' when importing tiff.
write_stars(summer_median_temp_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.tiff")

# I initially saved as RData, like this, but ran into problems later when trying to bind stars objects using c.stars(). It appears RData becomes corrutpt in a sense, over time, perhaps due to some time stamp. If I recreated the object in the global environment and saved as RDatam I could read it in and work with it, but the next day it would not work again.
#saveRDS(summer_median_temp_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.RData")
```

@@

Pivot and turn dimension into attributes:

```{r, eval=T}
summer_median_temp_ref_long <- summer_median_temp_ref %>% split("band")
```

The attribute name is *mean*. We can change this to something more meaningful.

```{r, eval=T}
summer_median_temp_ref_long <- setNames(summer_median_temp_ref_long, c("reference_upper", "sd"))
```

@@

#### Combine

@@

```{r}
y_var <- summer_median_temp %>%
  split(3) %>%
  c(summer_median_temp_ref_long)
```

### Normalise variable

@@

```{r, eval=F}
# Mutate

# The break point scaling is actually not needed here, since 
# having the lower ref value to be 5 sd implies that the threshold is
# 2 sd in a linear scaling.

@@

Writing to file, this time as RData (could use rds) because tiff don't allow multiple attributes, and because we don't need to use `c.stars()` in this object.

```{r, eval=F}
gc()
saveRDS(y_var_norm, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_normalised.RData")
```

@@

The *real* indicator values should be means over 5 year periods. Calculating a running mean for all time steps is too time consuming. Therefore, the scaled, regional indicator values will be calculated for distinct time steps. The time series can perhaps best be presented with yearly resolution.

### Plot time series as line graphs

We could also plot a time series as a series of maps, but then we need to average over these 5 year time periods first. We can get to that later.

To plot time series I first need to do a spatial aggregation. One option is to spatially aggregate the stars object directly, using the `reg` data set:

```{r, eval=F}
temp <- y_var_norm[1,,] %>%
  aggregate(reg, mean, na.rm=T)
plot(temp)
temp2 <- temp %>%
  st_as_sf() # too slow
```

However, this does no preserve the region name, and a subsequent `st_intersection` taks a little while and gives some boundary problems.

I can also convert pixels to points and then do the intersection.

```{r, eval=F}
temp_points <- y_var_norm[1,,] %>%
  st_as_sf(as_points = TRUE, merge = FALSE)
temp_points2 <- temp_points %>%
  st_intersection(reg) " too slow"
```

But this also took too long.

I can perhaps aggregate, and then turn to points and intersect, but that did not work perfect either.

I can rasterize the regions data set, turn it into a coded dimension and use st_apply, but that is tedious.

Let me try using `exact_extract`, as an alternative to `aggregate`. This requires us to turn to the `terra` package for a bit.

@@

That was fast!

I should also get the sd:

@@

```{r, include=F}
regional_means <- readRDS("temp/regional_means.rds")
regional_sd <- readRDS("temp/regional_sd.rds")
```

@@

#Adding the spatial sd
(
regional_means_long <- regional_sd %>%
  select(!all_of(div)) %>%
  pivot_longer(!region) %>%
  separate(name, into=c("type", "year")) %>%
  pivot_wider(names_from = type) %>%
  rename(i_sd = i,
         v_sd = v) %>%
  left_join(regional_means_long, by=join_by(region, year))
)
```

```{r, fig.cap="Times series for median summer temperature centered on the median value during the reference period. The reference period is indicated with a thick horizontal line. Dottet horisontal lines are 2 sd units for the reference period."}
regOrder = c("Østlandet","Sørlandet","Vestlandet","Midt-Norge","Nord-Norge")

regional_means_long %>%
  mutate(col = if_else(diff>0, "1", "2")) %>%
  ggplot(aes(x = as.numeric(year), 
           y = diff, fill = col))+
  geom_bar(stat="identity")+
  geom_hline(aes(yintercept = threshold_low_centered),
        linetype=2)+
  geom_hline(aes(yintercept = threshold_high_centered),
        linetype=2)+
  geom_segment(x = 1961, xend=1990,
               y = 0, yend = 0,
               linewidth=2)+
  scale_fill_hue(l=70, c=60)+
  theme_bw(base_size = 12)+
  ylab("Sommertempratur\navvik fra 1961-1990")+
  xlab("")+
  guides(fill="none")+
  facet_wrap( .~ factor(region, levels = regOrder),
              ncol=3,
              scales = "free_y")
```

Then we can take the mean over the last 5 years and add to a spatial representation.

```{r}
(
i_aggregatedToPeriods <- regional_means_long %>%
  mutate(year = as.numeric(year)) %>%
  mutate(period = case_when(
    year %between% c(2018, 2022) ~ "2018-2022",
    year %between% c(2013, 2017) ~ "2013-2017",
    year %between% c(2008, 2012) ~ "2008-2012",
    year %between% c(2003, 2007) ~ "2003-2007",
    .default = "pre 2003"
  )) %>%
  mutate(period_rank = case_when(
   period == "2018-2022" ~ 5,
   period == "2013-2017" ~ 4,
   period == "2008-2012" ~ 3,
   period == "2003-2007" ~ 2,
    .default = 1
  )) %>%
  group_by(region, period, period_rank) %>%
  summarise(indicator = mean(i),
            spatial_sd = sqrt(sum(i_sd^2))/length(i_sd)
            #,
            #spatial_sd_mean = mean(i_sd),
            #spatial_sd2_max = max(i_sd),
            #spatial_sd2_length = length(i_sd)
            )
)
```

```{r, fig.cap="Scaled indicator values, aggregated over 5 year intervals. Errors represent spatial variation within regions and across years."}
labs <- unique(i_aggregatedToPeriods$period[order(i_aggregatedToPeriods$period_rank)])

i_aggregatedToPeriods %>%
  ggplot(aes(x = period_rank, 
             y = indicator,
             colour=region))+
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin=indicator-spatial_sd, 
                    ymax=indicator+spatial_sd), 
                    width=.2,
                 position=position_dodge(0.2)) +
  theme_bw(base_size = 12)+
  scale_x_continuous(breaks = 1:5,
                     labels = labs)+
  labs(x = "", y = "indikatorverdi")
```

Finally, I can add these values to the sp object.

```{r}
reg2 <- reg %>%
  left_join(i_aggregatedToPeriods[i_aggregatedToPeriods$period_rank==5,], by=join_by(region))
names(reg2)
```

```{r, fig.cap="Summer tempreature indicator values for five accounting areas in Norway. SD is the spatial variation."}
myCol <- "RdYlGn"
myCol2 <- "-RdYlGn"

@@

tm_inset <- tm_shape(reg2)+
  tm_polygons(col="spatial_sd",
              title="SD",
              palette = myCol2,
              style="cont")+
  tm_layout(legend.format = list(digits=2))

>>>>

# Median Summer Temperature {#median-summer-temp}

@@

```{r, echo=F}
Ecosystem <- "All" 
Egenskap  <- "Abiotiske egenskaper"
ECT       <- "Physical state characteristics" 
Contact   <- "Anders L. Kolstad" 
```

@@

This chapters describes the workflow for an indicator describing the median summer temperature. For a more comprehensive documentation on the development of the workflow itself, see [here](#climate-indicators-explained). The data comes from interpolated climate surfaces from [SeNorge](https://senorge.no/) which contain one 1x1km raster for each day since 1957 to present. The reference levels are extracted from the time period 1961-1990 (a temporal reference condition).

**Workflow**

1.  Collate variable data series, ending up with one raster per year after aggregating across days within a year

2.  Calculate spatially explicit reference values by aggregating across the years 1961-1990. The upper reference value is equal to the median value of this period, and the lower reference values (two-sided) is equal to 5 SD units. The indicator value is the mean over a five year period.

3.  Calculate indicator values through spatially explicit rescaling based on the reference values

4.  Mask by ecosystem type (*This step is not yet done as we do not have ready available ecosystem maps*)

5.  Aggregate in space (to accounting areas) and take the mean over a five year period to get final indicator values

6.  Make trend figure and spatially aggregated maps

@@

We are interested in _tm_, which is temperatur in celcius. For some reason the same variable is called _tg_ in the data itself.

@@

```{r, warning=F}
reg <- st_intersection(reg, nor)
```

```{r, fig.cap="Five accounting areas (regions) in Norway."}
tm_shape(reg) +
  tm_polygons(col="region")

@@

6.  Make trend figure and spatially aggregated maps

### Step 1 - temporal aggregation within a year

@@

This takes about 20 sec per file/year, or 22 min on total. That is not too bad. About 6000 rasters are read into memory. Here's a test for the effect of splitting over more cores.

```{r, eval=F}
write_stars(real_temp_summer, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/real_temp_summer.tiff")
```

```{r, include=F}
real_temp_summer <- read_stars(paste0(pData, "climate_indicators/aggregated_climate_time_series/real_temp_summer.tiff"),
                                 proxy=F)
```

Note that GTiff automatically renames the third dimension *band* and also renames the attribute. I can rename them.

```{r}
summer_median_temp <- real_temp_summer %>% 
  st_set_dimensions(names = c("x", "y", "v_YEAR")) %>%
  setNames("temperature")
```

```{r, fig.cap="Showing three random slices of the year dimension."}
ggplot()+
  geom_stars(data = summer_median_temp[,,,c(1,11,66)], downsample = c(10, 10, 0)) +
  coord_equal() +
  theme_void() +
  scale_fill_viridis_c(option = "D") +  
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0))+
  facet_wrap(~v_YEAR)
```

### Step 2 - calculate reference values

We need to first to define a reference period and then to subset our data `summer_median_temp`.

First we need to rename our dimension values and turn them back into dates.

@@

```{r, eval=F}
write_stars(summer_median_temp_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.tiff")
```

@@

Pivot and turn dimension into attributes, and rename attributes:

```{r, eval=T}
summer_median_temp_ref_long <- summer_median_temp_ref %>% 
  split("band") %>%
  setNames(c("reference_upper", "sd"))
```

@@

@@

```{r}
y_var <- summer_median_temp %>%
  split("v_YEAR") %>%
  c(summer_median_temp_ref_long)
```

### Step 3 - normalise variable

@@

```{r, eval=F}
# The break point scaling is actually not needed here, since 
# having the lower ref value to be 5 sd implies that the threshold is
# 2 sd in a linear scaling.

@@

```{r, eval=F}
gc()
saveRDS(y_var_norm, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_normalised.RData")

# Tiff dont allow for multiple attributes:
#write_stars(y_var_norm, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_normalised.tiff")
```

@@

The *real* indicator values should be means over 5 year periods. Calculating a running mean for all time steps is too time consuming. Therefore, the scaled, regional indicator values will be calculated for distinct time steps. The time series can perhaps still be presented with yearly resolution.

### Step 4 - Mask with ecosystem delineation map

This step we simply ignore for now. It should be relatively easy to do, when we have the maps.

### Step 5 - Make figures

To plot time series I first need to do a spatial aggregation.

@@

I could also get the sd like this, if I wanted to base the indicator uncertainty on a measure of spatial variation:

@@

```{r, include=F}
regional_means <- readRDS("temp/regional_means.rds")
#regional_sd <- readRDS("temp/regional_sd.rds")
```

@@

#Adding the spatial sd
#(
#regional_means_long <- regional_sd %>%
#  select(!all_of(div)) %>%
#  pivot_longer(!region) %>%
#  separate(name, into=c("type", "year")) %>%
#  pivot_wider(names_from = type) %>%
#  rename(i_sd = i,
#         v_sd = v) %>%
#  left_join(regional_means_long, by=join_by(region, year))
#)
```

```{r summer-temp-time-series, fig.cap="Times series for median summer temperature centered on the median value during the reference period. The reference period is indicated with a thick horizontal line. Dottet horisontal lines are 2 sd units for the reference period."}
regOrder = c("Østlandet","Sørlandet","Vestlandet","Midt-Norge","Nord-Norge")

regional_means_long %>%
  mutate(col = if_else(diff>0, "1", "2")) %>%
  ggplot(aes(x = as.numeric(year), 
           y = diff, fill = col))+
  geom_bar(stat="identity")+
  geom_hline(aes(yintercept = threshold_low_centered),
        linetype=2)+
  geom_hline(aes(yintercept = threshold_high_centered),
        linetype=2)+
  geom_segment(x = 1961, xend=1990,
               y = 0, yend = 0,
               linewidth=2)+
  scale_fill_hue(l=70, c=60)+
  theme_bw(base_size = 12)+
  ylab("Sommertemperatur\navvik fra 1961-1990")+
  xlab("")+
  guides(fill="none")+
  facet_wrap( .~ factor(region, levels = regOrder),
              ncol=3,
              scales = "free_y")
```

Then we can take the mean and sd over the last 5 years and add to a spatial representation.

```{r}
(
i_aggregatedToPeriods <- regional_means_long %>%
  mutate(year = as.numeric(year)) %>%
  mutate(period = case_when(
    year %between% c(2018, 2022) ~ "2018-2022",
    year %between% c(2013, 2017) ~ "2013-2017",
    year %between% c(2008, 2012) ~ "2008-2012",
    year %between% c(2003, 2007) ~ "2003-2007",
    .default = "pre 2003"
  )) %>%
  mutate(period_rank = case_when(
   period == "2018-2022" ~ 5,
   period == "2013-2017" ~ 4,
   period == "2008-2012" ~ 3,
   period == "2003-2007" ~ 2,
    .default = 1
  )) %>%
  group_by(region, period, period_rank) %>%
  summarise(indicator = mean(i),
            sd = sd(i)
            # If I inluded a spatial measure for the uncertainty, here is how I would carry the errors:
            #spatial_sd = sqrt(sum(i_sd^2))/length(i_sd)
            )
)
```

```{r summer-temp-time-periods, fig.cap="Scaled indicator values, aggregated over 5 year intervals. Errors represent temporal variation (standard errors) within regions and across 5 years."}
labs <- unique(i_aggregatedToPeriods$period[order(i_aggregatedToPeriods$period_rank)])

i_aggregatedToPeriods %>%
  ggplot(aes(x = period_rank, 
             y = indicator,
             colour=region))+
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin=indicator-sd, 
                    ymax=indicator+sd), 
                    width=.2,
                 position=position_dodge(0.2)) +
  theme_bw(base_size = 12)+
  scale_x_continuous(breaks = 1:5,
                     labels = labs)+
  labs(x = "", y = "indikatorverdi")
```

Finally, I can add these values to the sp object with the accounting areas.

```{r}
reg2 <- reg %>%
  left_join(i_aggregatedToPeriods[i_aggregatedToPeriods$period_rank==5,], by=join_by(region))
```

```{r wall-to-wall-summer-temp-indicator, fig.cap="Summer tempreature indicator values for five accounting areas in Norway. SD is the spatial variation."}
myCol <- "RdYlGn"
myCol2 <- "-RdYlGn"

@@

tm_inset <- tm_shape(reg2)+
  tm_polygons(col="sd",
              title="SD",
              palette = myCol2,
              style="cont")+
  tm_layout(legend.format = list(digits=2))
